%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FILE         : Linear-Algebra.tex
% AUTHOR       : (C) Copyright 2010 by Peter C. Chapin
% SUBJECT      : Quick primer on the basics of linear algebra.
%
% To Do:
%
% + See the comments in the text for ideas about places where this document could be improved.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%+++++++++++++++++++++++++++++++++
% Preamble and global declarations
%+++++++++++++++++++++++++++++++++
\documentclass{article}
\setlength{\parindent}{0em}
\setlength{\parskip}{1.75ex plus0.5ex minus0.5ex}

\usepackage{hyperref}

% Useful commands to streamline my typing.
\newcommand{\abs}[1]{|#1|}
\newcommand{\norm}[1]{\|#1\|}

\newtheorem{theorem}{Theorem}[section]

%++++++++++++++++++++
% The document itself
%++++++++++++++++++++
\begin{document}

% Nice command for formatting new terms.
\newcommand{\newterm}[1]{\emph{#1}}

%-----------------------
% Title page information
%-----------------------
\title{Linear Algebra}
\author{Peter C. Chapin}
\date{February 6, 2010}
\maketitle

\tableofcontents

\section*{Legal}
\label{sec:legal}

\textit{Permission is granted to copy, distribute and/or modify this document under the terms of
  the GNU Free Documentation License, Version 1.1 or any later version published by the Free
  Software Foundation; with no Invariant Sections, with no Front-Cover Texts, and with no
  Back-Cover Texts. A copy of the license is included in the file \texttt{GFDL.txt} distributed
  with the \LaTeX\ source of this document.}

\section{Introduction}
\label{sec:intro}

This document is a quick primer on linear algebra. Its intended audience is senior computer
engineering technology students at Vermont Technical College. This document assumes the reader
has a reasonable mathematical background, largely calculus based, but no prior exposure to the
topics of linear algebra at all.

At the end of each major section below you will find some exercies that will encourage you to
practice with the concepts presented in that section. In addition you will find some notes on
how to use Octave to experiment with the topics of the section. Octave is an open source program
that excels at doing matrix computations of various kinds (as well as many other mathematical
things). It is similar to the commerical program MATLAB and even uses a largely compatible
syntax. You may also be able to do a number of interesting matrix computations on your
calculator. However, Octave will probably have greater capacity, higher performance, and more
functionality than your calculator. Regardless of what your calculator can or can't do, I
recommend that you spend some time getting familiar with Octave.

\input{concept.tex}
\input{operations.tex}
\input{linear-transform.tex}
\input{inverse.tex}
\input{norms.tex}
\input{equations.tex}
\input{gaussian.tex}
\input{eigen.tex}

\end{document}
